---
title: 'Mortality Modelling with R'
author: |
  | Andrés M. Villegas
  | \textcolor{white}{white}
  | \footnotesize School of Risk and Actuarial Studies, UNSW Sydney
  | \textcolor{white}{white}
  | \textcolor{white}{white}
  | \includegraphics[width = 6cm]{logocepar.png}
  | \textcolor{white}{white}
  | \textcolor{white}{white}
  | 15 January 2020, Living to 100 Symposium
  | Orlando, Florida
output: beamer_presentation
geometry: margin=1cm
template: defaultTop.tex
header-includes:
    - \definecolor{CEPAR}{RGB}{192, 8, 31}  
    - \setbeamercolor{structure}{fg=CEPAR}
    - \widowpenalties 1 150
    - \usepackage[orientation=landscape,size=custom,width=16,height=9,scale=0.5,debug]{beamerposter}
    - \DeclareMathOperator{\logit}{logit}
    - \newcommand{\ei}{\end{itemize}}
    - \newcommand{\bi}{\begin{itemize}}
    - \newcommand{\red}[1]{{\color{CEPAR} {#1}}}
    - \newcommand{\ra}{\red{$\leadsto$}\ }
    - \usepackage{booktabs}
    - \usepackage{animate}
includes:
  before_body: doc_prefix.txt
  in_header: header.txt
bibliography: bibliography.bib
nocite: | 
  @Brouhns2005, @Koissi2006
---


```{r lineWidthHook, eval=TRUE,echo = FALSE}
#Hook to control the line width in ouputs
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```


----

\vspace{3cm}

\centering


\text{{\Huge\color{CEPAR} Slides Available at:}}

\LARGE

https://github.com/amvillegas/LT100_2020

\vspace{3cm}


# Aims

Gain familiarity with use of the `R` packages **demography**, **StMoMo** and **lifecontingencies** to:

- Obtain mortality data from Human Mortality Database
- Fit stochastic mortality models (including Lee-Carter model)
- Compare goodness of fit of mortality models
- Forecast future mortality rates (deterministic and using Monte Carlo simulations)
- Compute demographic and actuarial quantities

# Agenda

- Motivation 
- Review of mortality modelling metodologies
- Mortality modelling with `R`
    - Downloading mortality data with **demography**
    - Fitting, comparing and projecting mortality model with **StMoMo**
    - Turning mortality projections into demographic calculations with **lifecontingencies**
- Conclusions and Summary
- Preview of recent **StMoMo** developments (if time permits!)


----

\vspace{3cm}
\begin{center}
\text{{\Huge\color{CEPAR} Motivation and Review}}

\text{{\Huge\color{CEPAR}  of Mortality Modelling}}
\end{center}
\vspace{3cm}


# Recent trends in mortality and life expectancy 

```{r mortalitytrends, eval=TRUE,  fig.width=12, fig.height=3.5, out.width="14.5cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE}
library(wpp2017)
library(tidyverse)
library(ggplot2)
library(cowplot)


data(e0F)
data(mxF)

LEDf <- e0F %>%  gather("year", "LE0", 3:15) %>% 
  rename(Region = name)

mxDf <- mxF %>%  gather("year", "LE0", 4:33) %>% 
  rename(Region = name)



countries <- c("WORLD", "ASIA", "LATIN AMERICA AND THE CARIBBEAN", "AFRICA", 
               "EUROPE", "NORTHERN AMERICA", "OCEANIA", "Australia")
# years <- paste(seq(1950,2045,5),seq(1955,2050,5), sep = "-")
years <- paste(seq(1950,2010,5),seq(1955,2015,5), sep = "-")

plotLE <- ggplot(filter(LEDf, Region %in% countries & year %in% years)) +
  geom_line(aes(x=year, y = LE0, group = Region, colour = Region)) +
  geom_point(aes(x=year, y = LE0, group = Region, colour = Region,
                 shape = Region)) +  theme_bw() + 
  labs(title = "Period Life Expectancy at brith (Females)", y = "Life Expectancy (years)",
       x = NULL) +
  theme(axis.text.x=element_text(angle=45,hjust=1)) 


plotMx <- ggplot(filter(mxDf, Region %in% countries & year %in% years & age == 60)) +
  geom_line(aes(x=year, y = LE0, group = Region, colour = Region)) +
  geom_point(aes(x=year, y = LE0, group = Region, colour = Region,
                 shape = Region)) +  theme_bw() + 
  labs(title = "Mortality rate at age 60-64 (Females)", y = "mortality rate",
       x = NULL) +
  theme(axis.text.x=element_text(angle=45,hjust=1)) + ylim(c(0, NA))


legend <- get_legend(plotMx)

plotMx <- plotMx +  theme(legend.position='none')
plotLE <- plotLE +  theme(legend.position='none')

plot_grid(plotMx, plotLE, legend, ncol=3)

```

\footnotesize Source: United Nations World Population Prospects 2017

\normalsize

- Good-news!
- Important social and financial implications for governments, insurers, individuals.
- Need to model and project these trends


# Mortality forecasting methodologies

A good overview of methodologies is given in the review papers by  @Booth2008, @Wong-Fupuy2004, @Pitacco2004 and in the  by @Pitacco2009

- Expert based

- Explanatory
    - Structural Modelling (Explanatory or Econometric).
    - Cause of death decomposition

- **Extrapolation**
    - Trend modelling


# A timeline of "recent" mortality modelling methodologies

\centering 
\includegraphics[width = \textwidth]{FiguresPreliminaries/Timeline}


# Advances in single population mortality modelling

* **Lee-Carter model** [@Lee1992] 
    + Add more bilinear age-period components [@Renshaw2003a]
    + Add a cohort effect [@Renshaw2006]
* Two factor **CBD model** [@Cairns2006a]
    + Add cohort effect, quadratic age term [@Cairns2009] 
    + Combine with features of the Lee-Carter [@Plat2009a]
* **Many more models** proposed in the literature (e.g. @Aro2011, @OHare2012, @Borger2013, @Alai2014b)

```{r downloadUSdata, eval=TRUE,echo = FALSE, message=FALSE, warning=FALSE, cache = TRUE}
library(StMoMo)
library(demography) 
USdata <- hmd.mx(country = "USA", username = "a.villegas@unsw.edu.au", password = "hmd")
Ext <- USdata$pop$male 
Dxt <- USdata$rate$male * Ext
Dxt <- Dxt[, as.character(1933:2017)]
Ext <- Ext[, as.character(1933:2017)]
years <- 1933:2017
```


# Lee-Carter model

\centering

```{r plotEWAni1933, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:1], type = "l", 
         lty = 1, col = 'black', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1933)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,1], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model

\centering

```{r plotEWAni1935, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:2], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1935)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,3], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1940, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:7], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1940)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,8], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1945, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:12], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1945)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,13], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1950, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:17], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1950)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,10], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotEWAni1955, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:22], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1955)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,23], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model

\centering

```{r plotEWAni1960, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:27], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1960)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,28], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model

\centering

```{r plotEWAni1965, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:32], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1965)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,33], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1970, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:37], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1970)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,38], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1975, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:42], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1975)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,43], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1980, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:47], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1980)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,48], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model

\centering

```{r plotEWAni1985, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:52], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1985)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,53], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni1990, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:57], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1990)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,58], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotEWAni1995, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:62], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1995)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,63], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni2000, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:67], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (2000)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,68], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotEWAni2005, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:72], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (2005)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,73], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni2010, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:77], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (2010)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,78], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotEWAni2015, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:82], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (2015)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,83], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$

# Lee-Carter model

\centering

```{r plotEWAni2017, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext)[,1:83], type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (2017)", ylim = c(-10, 0))
 lines(USdata$age, log(Dxt / Ext)[,84], lwd = 2)

```
$${\color{white}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model

\centering

```{r plotAPC1, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext), type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1933-2017)", ylim = c(-10, 0))

```
$${\color{black}\log \mu_{xt} \quad =}\quad {\color{white}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r estimateAPC, results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 APCfit <- fit(StMoMo(periodAgeFun = NULL), Dxt = Dxt, Ext = Ext, ages = USdata$age, years = years)
```

```{r plotAPC2, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext), type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1933-2017)", ylim = c(-10, 0))
 lines(APCfit$ages,APCfit$ax,col="black",lwd=2,lty=2)
```
$${\color{black}\log \mu_{xt} \quad =}\quad {\color{black}\alpha_x}  \quad{\color{white}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{white}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotAPC3, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext), type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1933-2017)", ylim = c(-10, 0))
  x <- seq(11,91,10)
  y0 <- apply(log(Dxt / Ext) [x,],1,max)
  y1 <- apply(log(Dxt / Ext) [x,],1,min)
  arrows(x, y0, x, y1, length = 0.15, col ="black",lwd=2)
```
$${\color{black}\log \mu_{xt} \quad =}\quad {\color{black}\alpha_x}  \quad{\color{black}+}\quad {\color{white}\sum_{i=1}^N} {\color{white} \beta_x^{{\color{white}(i)}}}{\color{black}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$


# Lee-Carter model

\centering

```{r plotAPC4, eval=TRUE,  fig.width=7, fig.height=4.5,out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
 matplot(USdata$age, log(Dxt / Ext), type = "l", 
         lty = 1, col = 'grey', xlab = "age", ylab = "log death rates", 
         main = "USA: male death rates (1933-2017)", ylim = c(-10, 0))
  x <- seq(11,91,10)
  y0 <- apply(log(Dxt / Ext) [x,],1,max)
  y1 <- apply(log(Dxt / Ext) [x,],1,min)
  arrows(x, y0, x, y1, length = 0.15, col ="darkgrey",lwd=2)
  
  x <- c(11,61)
  y0 <- apply(log(Dxt / Ext) [x,],1,max)
  y1 <- apply(log(Dxt / Ext) [x,],1,min)
  arrows(x, y0, x, y1, length = 0.15, col ="black",lwd=2)
  
  
```
$${\color{black}\log \mu_{xt} \quad =}\quad {\color{black}\alpha_x}  \quad{\color{black}+}\quad {\color{white}\sum_{i=1}^N} {\color{black} \beta_x^{{\color{white}(i)}}}{\color{black}\kappa_t}^{{\color{white}(i)}} \quad{\color{white}+}\quad {\color{white}\beta_x^{(0)}}{\color{white}\gamma_{t-x}}$$



# Lee-Carter model 

\centering


```{r plotLC, eval=TRUE,  fig.width=8*0.85, fig.height=6*0.85,out.width="10.5cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE, dependson=c("downloadUSdata")}
USmale <- StMoMoData(USdata, series = "male")
ages.fit <- 0:100
years.fit <- 1968:2017

#Fit all models         
LCfit <- fit(lc(), data = USmale, ages.fit = ages.fit,  
             years.fit = years.fit) 
plot(LCfit)  
```


#Cairns-Blake-Dowd model 

\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-2-1} 

 
\end{column}
\end{columns}



 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-3-1} 

 




#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-4-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-5-1} 

 




#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-6-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-7-1} 

 




#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-8-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-9-1} 






#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-10-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-11-1} 

 



#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-12-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-13-1} 

 



#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-14-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-15-1} 

 



#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-16-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-17-1} 

 




#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-18-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-19-1} 

 





#Cairns-Blake-Dowd model 
\begin{columns}
\begin{column}{0.4\textwidth}
\[\logit q_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} \]
\end{column}
\begin{column}{0.5\textwidth}
 
 
\includegraphics[width=4.5cm]{FigureCBD/unnamed-chunk-20-1} 

 
\end{column}
\end{columns}

 
 
\centering \includegraphics[width=9cm]{FigureCBD/unnamed-chunk-21-1} 


# Other Stochastic Mortality Models

```{r tab1, echo=FALSE, message=FALSE,warning=FALSE, cache=TRUE, autodep=TRUE}
library(kableExtra)
table <- data.frame("Name" = c("LC", "LC2", "RH", "APC", "CBD", "M7", "sPLAT", "cPLAT"), "Form" = c('$\\alpha_x + \\beta_x^{(1)} \\kappa_t^{(1)}$', '$\\alpha_x + \\beta_x^{(1)} \\kappa_t^{(1)} + \\beta_x^{(2)} \\kappa_t^{(2)}$', '$\\alpha_x + \\beta_x^{(1)} \\kappa_t^{(1)} + \\beta_x^{(0)}\\gamma_{c}$', '$\\alpha_x + \\kappa_t^{(1)} + \\gamma_{c}$', '$\\kappa_t^{(1)} + (x-\\bar{x}) \\kappa_t^{(2)}$', '$\\kappa_t^{(1)} + (x-\\bar{x})\\kappa_t^{(2)} + ((x-\\bar{x})^2 - \\sigma_x^2) \\kappa_t^{(3)} + \\gamma_{c}$', '$\\alpha_x + \\kappa_t^{(1)} + (\\bar{x}-x) \\kappa_t^{(2)} + \\gamma_{c}$', '$\\alpha_x + \\kappa_t^{(1)} + (\\bar{x}-x) \\kappa_t^{(2)} + (\\bar{x}-x)^+ \\kappa_t^{(3)} + \\gamma_{c}$'), "Parameters" = c('$2n_a + n_y$', '$3n_a + 2n_y$', '$3n_a + n_y + n_c$', '$n_a + n_y + n_c$', '$2n_y$', '$3n_y + n_c$', '$n_a + 2n_y + n_c$', '$n_a + 3n_y + n_c$'))
kable(table, "latex", align=c("c","l","l"), booktabs=TRUE, escape = F, linesep = "") %>%
kable_styling(latex_options = c("hold_position")) %>%
row_spec(0, align = "c")

```

#Generalised Age-Period-Cohort stochastic mortality models

Recent research has proposed a unifying framework discrete stochastic mortality models

-  General Age-Period-Cohort model structure [@Hunt2015b]
-  Generalised (non-)linear model [@Currie2014]
- R Implementation of GAPC models [@Villegas2018a]

#Generalised Age-Period-Cohort stochastic mortality models
\begin{enumerate}
  \item \red{Random Component:}
\[D_{xt}\sim \text{Poisson}(E^c_{xt}\mu_{xt}) \quad \text{or} \quad D_{xt}\sim \text{Binomial}(E_{xt}, q_{xt})\]
  
  \item \red{Systematic Component:}
  \[ \eta_{xt} \quad=\quad \alpha_x \quad+\quad {\sum_{i=1}^N \beta_x^{(i)}\kappa_t^{(i)} \quad+\quad \beta_x^{(0)}\gamma_{t-x}}\]
  \bi
    \item Lee-Carter type \ra $\beta_x^{(i)}$, non-parametric 
    \item CBD type \ra $\beta_x^{(i)}\equiv f^{(i)}(x)$, pre-specified parametric function 
  \ei
  
  \item \red{Link Function: }
    \[g\left(\mathbb{E}\left(\frac{D_{xt}}{E_{xt}}\right)\right) = \eta_{xt}\]
    \bi
      \item log-Poisson: $\eta_{xt} = \log \mu_{xt}$
      \item logit-Binomial: $\eta_{xt} = \logit q_{xt}$
    \ei
\end{enumerate}



#Generalised Age-Period-Cohort stochastic mortality models
\begin{enumerate}
\setcounter{enumi}{3}
  \item \red{Set of parameter constraints:}
  \bi

    \item Need parameters constraints to ensure identifiability
\ei

\item \red{Forecasting and simulation}
\bi
\item {Period indexes}: Multivariate random walk with drift
\begin{equation*}
\boldsymbol{\kappa}_t = \boldsymbol{\delta} + \boldsymbol{\kappa}_{t-1} + \boldsymbol{\xi}_t^\kappa, \qquad \boldsymbol{\kappa}_t = 
\left(\!
    \begin{array}{c}
      \kappa_t^{(1)} \\
      \vdots\\
  		\kappa_t^{(N)}
    \end{array}
  \!\right), \qquad \boldsymbol{\xi}_t^\kappa\sim N(\mathbf{0},\Sigma),
\end{equation*}

\item {Cohort effect}: ARIMA$(p,q,d)$ with drift 
\begin{equation*}
\Delta^d\gamma_c =  \delta_0 + \phi_1\Delta^d\gamma_{c-1} +\cdots+ \phi_p\Delta^d\gamma_{c-p}+\epsilon_{c}+\delta_1\epsilon_{c-1}+\cdots+\delta_q\epsilon_{c-q}
\end{equation*}
\ei
\end{enumerate}

<!-- \fbox{ -->
<!-- \resizebox{\textwidth}{!}{ -->
<!-- \begin{minipage}{1\textwidth} -->
<!-- GAPC models can be implemented with the R package StMoMo  (\url{http://cran.r-project.org/package=StMoMo})  -->
<!-- \end{minipage}}} -->


----

\vspace{3cm}
\begin{center}
\text{{\Huge\color{CEPAR}Mortality Modelling with \texttt{R}}}
\end{center}
\vspace{3cm}


# Mortality modelling in R

* **Demography** [@Hyndman2014] 
    * Download data from the Human Mortality Database
    * Lee-Carter model and several of its variants
* **ilc** [@Butt2014]
    * Lee-Carter with cohorts and Lee-Carter under a Poisson framework
* **Lifemetrics** (http://www.macs.hw.ac.uk/~andrewc/lifemetrics/ )
    * CBD and extensions
    * Lee-Carter with cohorts and Lee-Carter under a Poisson framework

* **StMoMo** [@Villegas2018a]
    * Implements Generalised Age-Period-Cohort Models


# Installation of `R` Packages

* To install the pacakges we use:
```{r, eval=FALSE}
install.packages("demography")
install.packages("StMoMo")
install.packages("lifecontingencies")
```

* To load within `R`:
```{r, eval=TRUE,echo = TRUE, message=FALSE, warning=FALSE}
library(demography)
library(StMoMo)
library(lifecontingencies)
```

# Downloading Data from the [Human Mortality Database](http://www.mortality.org/)

- Human Mortality Database (http://www.mortality.org/) contains population mortality data for 39 countries under consistent data protocol and in consistent format
    - Very useful for cross-country comparisons
    - Need to register to obtain username and password
- Subregional versions available for:
    - USA (https://usa.mortality.org/)
    - Japan (http://www.ipss.go.jp/p-toukei/JMD/index-en.asp)
    - Australia (http://demography.cass.anu.edu.au/research/australian-human-mortality-database) 
    - Canada (http://www.bdlc.umontreal.ca/CHMD/)
- For this example, we will use male USA data from 1950 to 2014


# Australian Data from [Human Mortality Database](http://www.mortality.org/)

```{r defineUSdata, echo=FALSE, eval=TRUE, cache=TRUE, results='hide'}
Ext <- USdata$pop$male 
Dxt <- USdata$rate$male * Ext
ages <- USdata$age     #0-110
years <- USdata$year   #1921-2011
```


```{r, eval=FALSE}
library(demography) 
USdata <- hmd.mx(country = "USA", username = "username",
                  password = "password")
```

```{r showDemogData, eval=TRUE, echo = TRUE, cache=TRUE, dependson = "defineUSdata"}
USdata
```


$$\text{{\color{white} x}}$$
$$\text{{\color{white} x}}$$
$$\text{{\color{white} x}}$$
$$\text{{\color{white} x}}$$
$$\text{{\color{white} x}}$$



# USA Data from [Human Mortality Database](http://www.mortality.org/)

```{r, eval=FALSE}
library(demography) 
USdata <- hmd.mx(country = "USA", username = "username",
                  password = "password")
```
\centering

\vspace{-0.5cm}

```{r plotDemogData, eval=TRUE,  fig.width=7, fig.height=4.15,out.width="9cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson=c("downloadUSdata")}
plot(USdata, series = "male")
```

# USA Data from [Human Mortality Database](http://www.mortality.org/)

```{r, eval=FALSE}
library(demography) 
USdata <- hmd.mx(country = "USA", username = "username",
                  password = "password")
```
\centering

\vspace{-0.5cm}

```{r plotDemogDataTime, eval=TRUE,  fig.width=7, fig.height=4.15,out.width="9cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson=c("downloadUSdata")}
plot(USdata, series = "male", plot.type = "time")
```

# **StMoMo**: **St**ochastic **Mo**rtality **Mo**delling
\centering
Who is MoMo?

\pause

\includegraphics[width = 11cm]{FiguresPreliminaries/momo}

$$\text{{\color{white} x}}$$

# **StMoMo**: **St**ochastic **Mo**rtality **Mo**delling
Who is MoMo?
\centering

\includegraphics[width = 11cm]{FiguresPreliminaries/Carnaval.jpg}

$$\text{{\color{white} x}}$$

# **StMoMo**: **St**ochastic **Mo**rtality **Mo**delling
\centering
Who is MoMo?

\includegraphics[width = 11cm]{FiguresPreliminaries/joselito.jpg}

# Overview of the structure of **StMoMo**

\centering

\includegraphics[width=12cm]{FiguresPreliminaries/StMoMoDiagram.png}

# GAPC stochastic mortality models with **StMoMo** 

Model         | Predictor ($\eta_{xt}$)
------------- | ----------------------------------------------------------
LC            | $\alpha_x + \beta^{(1)}_x \kappa^{(1)}_t$
APC           | $\alpha_x + \kappa^{(1)}_t+ \gamma_{t-x}$
CBD           | $\kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}$
M7            | $\kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)} + \left((x-\bar{x})^2-\hat{\sigma}_x^2\right)\kappa_t^{(3)} + \gamma_{t-x}$


* For consistency, all under a log-Poisson setting: 
$$D_{xt}\sim \mathrm{Poisson}(E^c_{xt}\mu_{xt})$$
$$\log \mu_{xt}=\eta_{xt}$$

# Model definition: Predefined functions for common models

. . .

```{r defineAll, eval=TRUE, cache=TRUE}
LC <- lc()
CBD <- cbd(link = "log")
APC <- apc()
M7 <- m7(link = "log")
```  

. . .

```{r showDefiniction, eval=TRUE, echo=FALSE, linewidth = 70}
LC 
CBD
APC
```  

```{r showDefiniction2, eval=TRUE, echo=FALSE, linewidth = 66}
M7 
```  

# Model fitting: Lee-Carter Example
\centering
```{r plotLCall, eval=TRUE,  fig.width=8, fig.height=5, out.width="9.5cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="defineAUSdata"}
USmale <- StMoMoData(USdata, series = "male")
LCfit <- fit(LC, data = USmale)
plot(LCfit)
```




# Model fitting: Concentrate on older ages
```{r fit1, eval=TRUE, cache=TRUE, results='hide', dependson=c("defineAll", "defineAUSdata"), warning=FALSE}
#Ages for fitting
ages.fit <- 60:89
years.fit <- 1968:2017

#Fit all models         
LCfit <- fit(LC, data=USmale, ages.fit=ages.fit,  
             years.fit=years.fit)
CBDfit <- fit(CBD, data=USmale, ages.fit=ages.fit,  
              years.fit=years.fit)
APCfit <- fit(APC, data=USmale, ages.fit=ages.fit,  
              years.fit=years.fit)
M7fit <- fit(M7, data=USmale, ages.fit=ages.fit,  
               years.fit=years.fit)
```


# Parameter estimates -- LC ($\log \mu_{xt} = \alpha_x + \beta^{(1)}_x \kappa^{(1)}_t$)
```{r plotLC2, eval=TRUE,  fig.width=8, fig.height=5.5, out.width="10cm", fig.align='center', results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson= c("defineAUSdata", "fit1")}
 plot(LCfit)
```


# Parameter estimates -- APC ($\log \mu_{xt} = \alpha_x + \kappa^{(1)}_t + \gamma_{t-x}$)
```{r plotAPC, eval=TRUE,  fig.width=8, fig.height=5.5, out.width="10cm", fig.align='center',  results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson= c("defineAUSdata", "fit1")}
 plot(APCfit, parametricbx = FALSE)
```


# Parameter estimates -- CBD ($\log \mu_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}$)
```{r plotCBD, eval=TRUE,  fig.width=8, fig.height=4, out.width="10cm", fig.align='center', results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson= c("defineAUSdata", "fit1")}
 plot(CBDfit, parametricbx = FALSE)
```

# Goodness-of-fit: Deviance Residuals


\begin{equation*}\label{eq:Residuals}
r_{xt}=\mathrm{sign}(d_{xt}-\hat{d}_{xt})\sqrt{\frac{\mathrm{dev}(x,t)}{\hat{\phi}}}
\end{equation*}

. . .


```{r resAll, eval=TRUE, echo=TRUE, cache=TRUE, dependson=c("fit1")}
#Compute residuals
LCres <- residuals(LCfit)
APCres <- residuals(APCfit)
CBDres <- residuals(CBDfit)
M7res <- residuals(M7fit)
```

# Goodness-of-fit: Residual heatmaps

```{r plotLCresShow, eval=FALSE}
plot(LCres, type = "colourmap", reslim = c(-3.5, 3.5))
``` 

\centering

```{r plotResHeatmaps1, fig.width=4, fig.height=3,out.width="4.5cm", echo=FALSE, cache=TRUE, dependson="resAll"}
par(mar=c(4.5,4,1,1))
plot(LCres, type = "colourmap", reslim = c(-3.5,3.5))
title("LC             ")
plot(APCres, type = "colourmap", reslim = c(-3.5,3.5))
title("APC            ")
``` 

```{r plotResHeatmaps2, fig.width=4, fig.height=3,out.width="4.5cm", echo=FALSE, cache=TRUE, dependson="resAll"}
par(mar=c(4.5,4,1,1))
plot(CBDres, type = "colourmap", reslim = c(-3.5,3.5))
title("CBD            ")
plot(M7res, type = "colourmap", reslim = c(-3.5,3.5))
title("M7             ")
``` 


# Goodness-of-fit: Residual scatterplots

```{r plotLCresScatter, eval=FALSE}
plot(CBDres, type = "scatter", reslim = c(-3.5, 3.5))
``` 

\centering

\vspace{-0.5cm}

```{r plotResScatter1, fig.width=10, fig.height=3,out.width="12cm", echo=FALSE, cache=TRUE, dependson="resAll"}
plot(CBDres, type = "scatter", reslim = c(-3.5,3.5))
title("CBD             ")
``` 


```{r plotResScatter2, fig.width=10, fig.height=3,out.width="12cm", echo=FALSE, cache=TRUE, dependson="resAll"}
plot(M7res, type = "scatter", reslim = c(-3.5,3.5))
title("M7              ")

``` 

# Goodness-of-fit vs Parsimony: AIC and BIC

\vspace{-0.5cm}

$$AIC = -2{\cal L} + 2 \times \nu \qquad BIC = -2{\cal L} +  \log K \times \nu $$

\vspace{-0.5cm}

. . .

```{r AIC, eval=TRUE, echo=TRUE, cache=TRUE, dependson=c("fit1")}
#Compute residuals
AIC(CBDfit)
BIC(CBDfit)
```

\vspace{-0.5cm}

```{r roundAIC, eval=TRUE, echo=FALSE, cache=FALSE, dependson=c("fit1")}
#Compute residuals
IC <- function(x) sprintf("%.0f", AIC(x)[1])
IC2 <- function(x) sprintf("%.0f", BIC(x)[1])
```

. . .

    
Criterion              |  LC              |  APC            |  CBD            | M7                  
---------------------- | ---------------- |---------------- |-----------------|--------------------- 
*AIC*                  | `r IC(LCfit )`   | `r IC(APCfit)`  | `r IC(CBDfit)`  | **`r IC(M7fit )`** 
*BIC*                  | `r IC2(LCfit )`  | `r IC2(APCfit)` | `r IC2(CBDfit)` | **`r IC2(M7fit )`**   

# Forecasting and simulation

* **Period indexes:** Multivariate random walk with drift
\begin{equation*}
\boldsymbol{\kappa}_t = \boldsymbol{\delta} + \boldsymbol{\kappa}_{t-1} + \boldsymbol{\xi}_t^\kappa, \qquad \boldsymbol{\kappa}_t = 
\left(\!
    \begin{array}{c}
      \kappa_t^{(1)} \\
      \vdots\\
  		\kappa_t^{(N)}
    \end{array}
  \!\right), \qquad \boldsymbol{\xi}_t^\kappa\sim N(\mathbf{0},\Sigma),
\end{equation*}

* **Cohort effect:** ARIMA$(p,q,d)$ with drift 
\begin{equation*}
\Delta^d\gamma_c =  \delta_0 + \phi_1\Delta^d\gamma_{c-1} +\cdots+ \phi_p\Delta^d\gamma_{c-p}+\epsilon_{c}+\delta_1\epsilon_{c-1}+\cdots+\delta_q\epsilon_{c-q}
\end{equation*}

# Forecasting


Model         | Model for $\gamma_{t-x}$
------------- | -------------
APC           | ARIMA$(1,1,0)$ with zero mean
M7            | ARIMA$(0,0,0)$ with zero mean

. . .

30-year ahead ($h=30$) central projections: period indexes, cohort index, and death rates probabilities:

```{r forAll, cache=TRUE, dependson=c("fit1")}
LCfor <- forecast(LCfit, h=30)
APCfor <- forecast(APCfit, h=30, gc.order = c(1,1,0), 
                   gc.include.constant = FALSE)
CBDfor <- forecast(CBDfit, h=30)
M7for <- forecast(M7fit, h=30, gc.order = c(0,0,0), 
                  gc.include.constant = FALSE)
``` 

# Forecasted period and cohort indexes
\centering
```{r plotForecastM7, eval=TRUE,  fig.width=8*1.0, fig.height=7*0.8, out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="forAll"}
plot(M7for, parametricbx = FALSE)
``` 

# Forecasted period and cohort indexes
\centering
```{r plotForecastAPC, eval=TRUE,  fig.width=8*1.0, fig.height=7*0.8, out.width="10cm", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="forAll"}
plot(APCfor, parametricbx = FALSE)
``` 


# Simulation

```{r simAll, cache=TRUE, dependson=c("fit1")}
LCsim <- simulate(LCfit, nsim=1000, h=30)
APCsim <- simulate(APCfit, nsim=1000, h=30, 
                   gc.order = c(1,1,0), 
                   gc.include.constant = FALSE)
CBDsim <- simulate(CBDfit, nsim=1000, h=30)
M7sim <- simulate(M7fit, nsim=1000, h=30, 
                  gc.order = c(0,0,0),
                  gc.include.constant = FALSE)
```


# Simulation 


```{r plotSimLC, fig.width=6, fig.height=4,out.width="6.5cm", echo=TRUE, cache=TRUE,fig.align='center', dependson="simAll"}
#Plot period index trajectories for the LC model
plot(LCfit$years, LCfit$kt[1,], type="l", xlim=c(1968,2047),
     ylim=c(-40,12), xlab="year", ylab="kt",
     main="Period index (LC)")
matlines(LCsim$kt.s$years, LCsim$kt.s$sim[1,,1:20], type="l", 
         lty=1)
```


# Fancharts


```{r plotLCFanShow, fig.width=6, fig.height=4,out.width="6.5cm", echo=TRUE, cache=TRUE,fig.align='center', dependson="simAll"}
library(fanplot)
mxt <- LCfit$Dxt/LCfit$Ext
plot(LCfit$years, mxt["65",], pch=20, xlim=c(1968,2047), 
     ylim=c(0.0045,0.04), log="y", xlab="year", ylab="m(65,t)")
fan(t(LCsim$rates["65",,]), start=2017, probs=c(50,75,90,97.5), 
  n.fan=4, ln=NULL, fan.col=colorRampPalette(c("black","white")))
```


# Fancharts
```{r fanchartFun, echo=FALSE, cache=TRUE}

rateFanChartFun<- function(GAPCfit, GAPCsim){
  probs = c(2.5, 10, 25, 50, 75, 90, 97.5)
  mxt <- LCfit$Dxt/LCfit$Ext
  #age 65
  plot(GAPCfit$years, mxt["65", ], xlim = c(1968, 2047), ylim = c(0.005, 0.2),
  xlab = "year", ylab = "mortality rate (log scale)",
  pch = 20, log = "y")
  fan(t(GAPCsim$rates["65", , ]), start = 2017, probs = probs, n.fan = 4,
  fan.col = colorRampPalette(c(rgb(80/255,81/255,84/255), "white")), ln = NULL)
  #age 75
  points(GAPCfit$years, mxt["75", ], pch = 20)
  fan(t(GAPCsim$rates["75", , ]), start = 2017, probs = probs, n.fan = 4,
  fan.col = colorRampPalette(c(rgb(0,170/255,176/255), "white")), ln = NULL)
  #age 75
  points(GAPCfit$years, mxt["85", ], pch = 20)
  fan(t(GAPCsim$rates["85", , ]), start = 2017, probs = probs, n.fan = 4,
  fan.col = colorRampPalette(c(rgb(192/255,8/255,31/255), "white")), ln = NULL)
  #labels
  text(1975, mxt[c("65", "75", "85"), "1995"],
  labels = c("x = 65", "x = 75", "x = 85"))
}
```

\centering

\vspace{-0.5cm}

```{r plotLC_APCfan, fig.width=3.8*0.9, fig.height=5*0.9,out.width="3.4cm", echo=FALSE, cache=TRUE, dependson=c("simAll", "fanchartFun")}
rateFanChartFun(LCfit, LCsim)
title("LC")
rateFanChartFun(APCfit, APCsim)
title("APC")
rateFanChartFun(CBDfit, CBDsim)
title("CBD")
rateFanChartFun(M7fit, M7sim)
title("M7")
```


# Obtaining projected life tables for a cohort

```{r LCcohort, echo=TRUE, cache=TRUE, dependson=c("forAll")}
chosen_cohort <- 1950

#observed rates for ages 0-59
hist_rates <- extractCohort(USmale$Dxt/USmale$Ext, 
                               cohort = chosen_cohort)[1:60]

#fitted historical rates for ages 60-67
lc_fit_rates <- extractCohort(fitted(LCfit, type = "rates"),
                               cohort = chosen_cohort)
#forcasted rates for ages 68-89
lc_for_rates <- extractCohort(LCfor$rates, 
                              cohort = chosen_cohort)
#all rates
lc_rates_1950 <- c(hist_rates,lc_fit_rates, lc_for_rates)
```



# Obtaining projected life tables for a cohort

```{r plotLCcohort, fig.width=5*0.9, fig.height=4*0.9,out.width="6.5cm", fig.align='center', echo=TRUE, cache=TRUE, dependson=c("LCcohort")}
plot(0:89, lc_rates_1950 ,type = "l", log = "y", xlab = "age",
     ylab = "m(x)", main = "Cohort 1950 mortality rate")
lines(68:89, lc_for_rates, col = "red")
```

# Computing life expectancies using **lifecontingencies**


```{r LE, echo=TRUE, cache=TRUE, dependson=c("LCcohort")}
#Transform mx to qx
lc_qx_1950 <- mx2qx(lc_rates_1950)
#Create lifetable
lc_lifetable_1950 <- probs2lifetable(probs = lc_qx_1950, 
                                     type = "qx",
                                     name = "LC-1950")
exn(lc_lifetable_1950,x=65)
```

```{r LEcohort, echo=FALSE, cache=TRUE, dependson=c("forAll", "LCcohort")}
cbd_rates_1950 <- c(hist_rates,
                   extractCohort(fitted(CBDfit, type = "rates"),
                               cohort = chosen_cohort),
                   extractCohort(CBDfor$rates, 
                              cohort = chosen_cohort))
apc_rates_1950 <- c(hist_rates,
                   extractCohort(fitted(APCfit, type = "rates"),
                               cohort = chosen_cohort),
                   extractCohort(APCfor$rates, 
                              cohort = chosen_cohort))
m7_rates_1950 <- c(hist_rates,
                   extractCohort(fitted(M7fit, type = "rates"),
                               cohort = chosen_cohort),
                   extractCohort(M7for$rates, 
                              cohort = chosen_cohort))
cbd_qx_1950 <- mx2qx(cbd_rates_1950)
cbd_lifetable_1950 <- probs2lifetable(probs = cbd_qx_1950, type = "qx")
apc_qx_1950 <- mx2qx(apc_rates_1950)
apc_lifetable_1950 <- probs2lifetable(probs = apc_qx_1950, type = "qx")
m7_qx_1950 <- mx2qx(m7_rates_1950)
m7_lifetable_1950 <- probs2lifetable(probs = m7_qx_1950, type = "qx")
```

. . . 


Model                  |  LC                              |  APC                              |  CBD                              | M7                  
---------------------- | -------------------------------- |---------------------------------- |----------------------------------- |---------------------------------- 
LE at age 65           | `r exn(lc_lifetable_1950,x=65)`  | `r exn(apc_lifetable_1950,x=65)`  | `r exn(cbd_lifetable_1950,x=65)`  | `r exn(m7_lifetable_1950,x=65)` 
LE at age 75           | `r exn(lc_lifetable_1950,x=75)`  | `r exn(apc_lifetable_1950,x=75)`  | `r exn(cbd_lifetable_1950,x=75)`  | `r exn(m7_lifetable_1950,x=75)`

# Summary

- This has been a whistle-stop introduction to fitting mortality models in `R`

- Useful example for using **StMoMo** in the package vignette at
https://cran.r-project.org/web/packages/StMoMo/vignettes/StMoMoVignette.pdf 
    - Also, references within this useful for understanding more about mortality models
    
-  Useful examples of integrating **StMoMo** with **lifecontingencies** at https://rdrr.io/cran/lifecontingencies/f/inst/doc/mortality_projection.pdf    

# Summary

* **StMoMo provides** easy implementation and comparison of a wide range of models making it useful for:
    * Actuaries analysing longevity risk $\leadsto$ Model risk
    * Use in the classroom

- Standard packages are starting point – however, to go beyond them you need to understand the principles behind their operation

# Work in progress

- Available in the development version and expected to be released in next six months  
    - Selection of models using cross-validation
    - Construction of models using regularisation techniques

* Sister packages **iMoMo** for mortality improvement rate modelling
    - Expected to be released in the next year
    
- Other development plans
    - Multipopulation models
    - Model combination 
    
----

\vspace{3cm}
\begin{center}
\text{{\Huge\color{CEPAR}Cross validation and regularisation of mortality models }}
\end{center}
\vspace{3cm}



# Mortality Models: Key research questions

1. What model features are desired for different applications?

\vspace{0.5cm}

2. Why do we only consider a fixed set of models?

\vspace{0.5cm}

3. How can we be confident we have selected the best model?

# Objective

Provide a comprehensive framework to **construct**, **select**, and **evaluate** discrete-time mortality models for forecasting applications, using various **statistical learning** and predictive analytics techniques.

- **Construction** based on regularisation techniques

- **Selection** based on cross-validation techniques


# Selection of mortality models using cross validation

\begin{figure}[h]
\centerline{\animategraphics[width=8.0cm, autoplay, loop]{2}{FiguresSL/cv}{4}{4}}
\end{figure}
\vspace{-1cm}

- \textbf{Training/Test Set}

- \textbf{Test Set Width}: Depending on forecasting Horizon

- \textbf{Metric}: MSE on the test set


# Selection of mortality models using cross validation

\begin{figure}[h]
\centerline{\animategraphics[width=8.0cm, autoplay, loop]{2}{FiguresSL/cv}{4}{16}}
\end{figure}
\vspace{-1cm}

- \textbf{Training/Test Set}

- \textbf{Test Set Width}: Depending on forecasting Horizon

- \textbf{Metric}: MSE on the test set

# Selection of mortality models using cross validation

```{r cv1, eval=TRUE, cache=TRUE, results='hide', dependson=c("defineAll", "defineAUSdata"), warning=FALSE}
# install.packages("devtools")
# devtools::install_github("amvillegas/StMoMo", 
#                            ref = "GroupLasso")

#CV for 1 year ahead     
LCcv1 <- cv.StMoMo(LC, h = 1, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
APCcv1 <- cv.StMoMo(APC, h = 1, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
CBDcv1 <- cv.StMoMo(CBD, h = 1, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
M7cv1 <- cv.StMoMo(M7, h = 1, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
```

# Selection of mortality models using cross validation

```{r cv10, eval=TRUE, cache=TRUE, results='hide', dependson=c("defineAll", "defineAUSdata"), warning=FALSE}

#CV for 10 year ahead     
LCcv10 <- cv.StMoMo(LC, h = 10, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
APCcv10 <- cv.StMoMo(APC, h = 10, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
CBDcv10 <- cv.StMoMo(CBD, h = 10, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
M7cv10 <- cv.StMoMo(M7, h = 10, data=USmale, ages.train=ages.fit,  
             years.train=years.fit, type = "logrates")
```

# Selection of mortality models using cross validation

```{r cvShow, eval=TRUE, cache=TRUE, dependson=c("cv10", "cv1"), warning=FALSE}
LCcv1$cv.mse
LCcv10$cv.mse
```

\vspace{-0.5cm}

```{r roundCV, eval=TRUE, echo=FALSE, cache=FALSE}
#Compute residuals
CV <- function(x) sprintf("%.4f", x$cv.mse)
```

. . .

    
Criterion              |  LC              |  APC                |  CBD            | M7                  
---------------------- | ---------------- |-------------------- |-----------------|--------------------- 
*AIC*                  | `r IC(LCfit )`   | `r IC(APCfit)`      | `r IC(CBDfit)`  | **`r IC(M7fit )`** 
*BIC*                  | `r IC2(LCfit )`  | `r IC2(APCfit)`     | `r IC2(CBDfit)` | **`r IC2(M7fit )`**   
*CV 1-year*            | `r CV(LCcv1 )`   | `r CV(APCcv1 )`     | `r CV(CBDcv1 )` | **`r CV(M7cv1 )`**   
*CV 10-year*           | `r CV(LCcv10)`   | **`r CV(APCcv10)`** | `r CV(CBDcv10)` | `r CV(M7cv10)`   


# Construction: "Formalised" Model-Building Framework 

We start with a \textbf{huge} model,

\begin{equation*} 
\ln(\mu_{x,t}) = \alpha_{x} + \sum_{i=1}^{B} f^{(i)}(x) \kappa_{t}^{(i)} + \gamma_{t-x},
\label{eq: GAPC Predictor Structure}
\end{equation*}

where the suite of basis functions $(f^{(i)}(x))$ included are:

\begin{equation} 
f^{(i)}(x) =
\begin{cases}
1, & \text{Unit }  \\
(x-\bar{x})^n & \text{Polynomial } \\
(x-n)^+ & \text{Call } \\
(n-x)^+ & \text{Put } \\
1_{x<n} & \text{Below } \\
1_{x>n} & \text{Above } 
\end{cases}
.
\end{equation}

# Construction: Our Model

\begin{figure}
    \centering
    \includegraphics[width=12.5cm, trim={0.0cm 0.0cm 0.0cm 0.0cm}, clip]{FiguresSL/functions_landscape.pdf}\\
\end{figure}

# Construction: GLM Representation
\begin{equation*}
\eta_{x,t} = \ln(\mu_{x,t}) = \alpha_x + \sum_{i=1}^B f^{(i)}(x) \kappa_t^{(i)} + \gamma_c,
\label{eq: Predictor}
\end{equation*}

can be expressed as a GLM [@Currie2014],
\begin{equation*}
\mathbf{\eta} = \mathbf{X \beta} = \sum_{j=0}^{B+1} \mathbf{X}_j \boldsymbol{\beta}_j, \qquad \mathbf{X} = [\mathbf{X_0}:\mathbf{X_{1}}:\mathbf{X_{2}}: ... : \mathbf{X_{B}}: \mathbf{X_{B+1}}],
\end{equation*}

where,
\vspace{-0.3cm}
\begin{equation*}
\boldsymbol{\beta}=\{\boldsymbol{\beta_i}\}_{i=0}^{B+1}, \quad \boldsymbol{\beta_0}=\{\alpha_x\}_{x=1}^{n_x}, \quad \boldsymbol{\beta_i}=\{\kappa_t^{(i)}\}_{t=1}^{n_t}, \quad \boldsymbol{\beta_{B+1}}=\{\gamma_c\}_{c=1}^{n_c}.
\end{equation*}

- Estimate parameters with group lasso using the R package **grpreg** [@Breheny2013]




````{r modelDef, echo=FALSE, warning=FALSE, cache=TRUE, message = FALSE}

strikes <- seq(65,85,5)
bModel <- StMoMo(link = "log-Gaussian", staticAgeFun = TRUE,
                 periodAgeFun = c("1", genPoly(1:5), genCall(strikes),
                                  genPut(strikes)),
                 cohortAgeFun = "1")

ktlabels <- c(substitute(paste("1", ""), list(i = 1)))

for(i in 1:5){
  ktlabels <- c(ktlabels, substitute(paste("(x-", xbar, ")"^{i}, ""), 
                                     list(i = i, xbar = mean(ages.fit))))
}
for(i in 1:length(strikes)){
  ktlabels <- c(ktlabels, substitute(paste("(x-", a, ")"^{i}, ""), 
                                     list(i = "+", a = strikes[i])))
}
for(i in 1:length(strikes)){
  ktlabels <- c(ktlabels, substitute(paste("(", a, "-x)"^{i}, ""), 
                                     list(i = "+", a = strikes[i])))
}
source("AnimationPlots.R")

```

````{r USAMale, echo=FALSE, warning=FALSE, cache=TRUE, dependson = c("modelDef", "defineAll", "defineAUSdata"), message = FALSE, results='hide' }

lambda <- exp(seq(log(0.03), log(0.0005), length.out = 50))
bMgrpfit <- grpfit(bModel, lambda = lambda, data = USmale, ages.fit = ages.fit,
                   years.fit = years.fit)

agesplot <- as.character(seq(60,85,5))
for(j in 1:length(lambda)){
  pdf(paste("./USAMale/parameters",j,".pdf",sep=""), height=4, width=6.5)
  plotParamPath(bMgrpfit, j, ktlabels)
  dev.off()
  
  pdf(paste("./USAMale/fit",j,".pdf",sep=""), height=4, width=6.5)
  plotFitPath(bMgrpfit, j, agesplot)
  dev.off()
}

```


# Construction: Estimated Parameters (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/parameters"}{1}{1}}
\end{figure}

# Construction: Estimated Parameters (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/parameters"}{1}{18}}
\end{figure}


# Construction: Estimated Parameters (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/parameters"}{18}{31}}
\end{figure}


# Construction: Estimated Parameters (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/parameters"}{31}{44}}
\end{figure}


# Construction: Estimated Parameters (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/parameters"}{44}{50}}
\end{figure}


# Construction: Fitted Values (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/fit"}{1}{1}}
\end{figure}

# Construction: Fitted Values (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/fit"}{1}{18}}
\end{figure}



# Construction: Fitted Values (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/fit"}{18}{31}}
\end{figure}



# Construction: Fitted Values (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/fit"}{31}{44}}
\end{figure}

# Construction: Fitted Values (USA Males)

\begin{figure}[h]
\centerline{\animategraphics[autoplay, width=13.0cm]{6}{"./USAMale/fit"}{44}{50}}
\end{figure}

#Implementation in StMoMo - Define model

\small

```{r defineModel, cache=TRUE, warning=FALSE, echo=TRUE, linewidth = 66}
#Create big model
strikes <- seq(65,85,5)
bModel <- StMoMo(link = "log-Gaussian", staticAgeFun = TRUE, 
                   periodAgeFun = c("1", genPoly(1:5), genCall(strikes),
                                    genPut(strikes)),
                   cohortAgeFun = "1")
bModel
```


#Implementation in StMoMo - Fit model with grouped penalised regularisation



```{r fitModel, cache=TRUE, warning=FALSE, echo=TRUE, linewidth = 66, dependson=c("defineModel", "loadData")}
#lambda grid
lambda <- exp(seq(log(0.03), log(0.0005), length.out = 50))
#Penalise all term
bMgrpfit <- grpfit(bModel, lambda = lambda, data = USmale,
                   ages.fit = ages.fit, years.fit = years.fit)

```

#Implementation in StMoMo - Parameter paths

```{r plotModel1, eval=TRUE,  fig.width=7.5*1.5, fig.height=3*1.5, out.width="100%", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="fitModel"}
 plot(extractStMoMo(bMgrpfit, 1), nCol = 3)
```

#Implementation in StMoMo - Parameter paths

```{r plotModel2, eval=TRUE,  fig.width=7.5*1.5, fig.height=3*1.5, out.width="100%", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="fitModel"}
 plot(extractStMoMo(bMgrpfit, 17), nCol = 3)
```

#Implementation in StMoMo - Parameter paths

```{r plotModel3, eval=TRUE,  fig.width=7.5*1.5, fig.height=3*1.5, out.width="100%", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="fitModel"}
 plot(extractStMoMo(bMgrpfit, 30), nCol = 3)
```

#Implementation in StMoMo - Parameter paths

```{r plotModel4, eval=TRUE,  fig.width=7.5*1.5, fig.height=3*1.5, out.width="100%", results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson="fitModel"}
 plot(extractStMoMo(bMgrpfit, 40), nCol = 5)
```

<!-- #Implementation in StMoMo - Select with crossvalidation -->

<!-- ```{r cvModel, cache=TRUE, warning=FALSE, echo=TRUE, linewidth = 66, dependson=c("defineModel", "loadData"), results='hide'} -->
<!-- #Penalise all term -->
<!-- bMgrpcv1 <- cv.grpStMoMo(bModel, h = 1, lambda = lambda, -->
<!--                          data = USmale, ages.train = ages.fit,  -->
<!--                          years.train = years.fit,  -->
<!--                          type = "logrates") -->
<!-- bMgrpcv10 <- cv.grpStMoMo(bModel, h = 10, lambda = lambda, -->
<!--                          data = USmale, ages.train = ages.fit,  -->
<!--                          years.train = years.fit,  -->
<!--                          type = "logrates") -->
<!-- ``` -->

<!-- #Implementation in StMoMo - Select with crossvalidation -->

<!-- ```{r cvBest, cache=TRUE, warning=FALSE, echo=TRUE, linewidth = 66, dependson=c("cvModel")} -->
<!-- #Best model for 1 year -->
<!-- bMgrpcv1$min -->
<!-- bMgrpcv1$lambda.min	 -->

<!-- #Best model for 10 year -->
<!-- bMgrpcv10$min -->
<!-- bMgrpcv10$lambda.min	 -->
<!-- ``` -->


----

\vspace{3cm}
\begin{center}
\text{{\Huge\color{CEPAR}Improvement rate modelling}}
\end{center}
\vspace{3cm}


# The APCi model

- Modelling improvement rates rather than mortality rates is becoming commong 
- Consider the APCi model which corresponds broadly to the  lates CMI projection approach:

$$-\log \frac{\mu_{xt}}{\mu_{x,t-1}} = \alpha_x  + \kappa_t + \gamma_{t-x}$$

# Implementation in **iMoMo**

```{r fitapci, eval=TRUE, cache=TRUE, dependson=c("defineAll", "defineAUSdata"), warning=FALSE}
library(iMoMo) #Not yet publicly available
#Define model
APCi <- apci()
APCi

#fit the model
APCifit <- fit(APCi, data=USmale, ages.fit=ages.fit,  
              years.fit=years.fit)
```

# Implementation in **iMoMo**

```{r plotAPCi, eval=TRUE,  fig.width=8, fig.height=5.5, out.width="10cm", fig.align='center',  results='hide', warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, dependson= c("defineAUSdata", "fitapci")}
 plot(APCifit, parametricbx = FALSE)
```



----
$$\text{{\color{white} x}}$$


http://cran.r-project.org/web/packages/StMoMo/
https://github.com/amvillegas
$$\text{{\color{white} x}}$$
\begin{center}
\text{{\Huge\textbf{Thank you!}}}
\end{center}
$$\text{{\color{white} x}}$$
a.villegas@unsw.edu.au

andresmauriciovillegas@gmail.com
\begin{flushright}
\includegraphics[width = 5cm]{logocepar.png}
\end{flushright}



# References {.allowframebreaks}
